Reduced plasticity and opportunity for selection in terrestrial ectotherm populations under climate change

### 2023-06-07 ms.qmd
Fixed p-value notation

The final dataset included a total of `r f <- taxa_sum |> dplyr::filter(habitat == "f") |> dplyr::summarise(spp = sum(spp), k = sum(n)); f[1,1]` freshwater (fishes = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Actinopterygii") |> dplyr::summarise(sp = sum(spp))`, Molluscs = `r taxa_sum |> dplyr::filter(habitat == "f" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`, Amphibians = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Amphibia") |> dplyr::summarise(sp = sum(spp))`, Reptiles = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))`, Arthropods = `r taxa_sum |> dplyr::filter(habitat == "f" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`, and a single Crustacean and Nematode species), `r m <- taxa_sum |> dplyr::filter(habitat == "m") |> dplyr::summarise(spp = sum(spp), k = sum(n)); m[1,1]` marine (fishes = `r taxa_sum |> dplyr::filter(habitat == "m" & class %in% c("Actinopterygii", "Agnatha","Chondrichthyes")) |> dplyr::summarise(sp = sum(spp))`, Annelids = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Annelida") |> dplyr::summarise(sp = sum(spp))`, Molluscs = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`, Echinoderms = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Echinodermata") |> dplyr::summarise(sp = sum(spp))`, Reptiles = `r taxa_sum |> dplyr::filter(habitat == "m" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))`, Arthropods = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`, and a single Crustacean and Cnidarian species), and `r t <- taxa_sum |> dplyr::filter(habitat == "t") |> dplyr::summarise(spp = sum(spp), k = sum(n)); t[1,1]` terrestrial species (Annelids = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Annelida") |> dplyr::summarise(sp = sum(spp))`, Molluscs = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`, Arthropods = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`, Reptiles = `r taxa_sum |> dplyr::filter(habitat == "t" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))` and Amphibians = `r taxa_sum |> dplyr::filter(habitat == "t" & class == "Amphibia") |> dplyr::summarise(sp = sum(spp))` along with a single Tardigrade species) (@fig-1 A). We had more data on acute thermal responses (n = `r trimmed <- data_long2 |> dplyr::filter(grepl("lnRR", name) & !is.na(effect_size) & !is.na(V) & !V == 0); sum1 = trimmed |> dplyr::group_by(name, type) |> dplyr::summarise(n = dplyr::n()) |> dplyr::filter(type =="acute"); sum(sum1$n)`) compared to thermal responses after an acclimation period (n = `r sum2 <- trimmed |> dplyr::group_by(name, type) |> dplyr::summarise(n = dplyr::n()) |> dplyr::filter(type =="acclim"); sum2$n`) because both acclimation temperatures had separate acute responses (@fig-1). While the two acute $lnRR_{Q_{10}}$ effect sizes did differ significantly from each other, on average (Acute responses were higher for animals acclimated to high temperatures -- $\beta$ = `r model1.s1$b[3]`, 95% CI: `r model1.s1$ci.lb[3]` to `r model1.s1$ci.ub[3]`, p `r p_value(model1.s1$pval[3])`), they were in the same direction and only differed by ~10%. As such, we averaged the two acute $lnRR_{Q_{10}}$ effect sizes in subsequent analyses.

@@

Overall, $lnRR_{Q_{10_{acclimation}}}$ and $lnRR_{Q_{10_{acute}}}$ differed by only `r mean(avg_change)`% across all habitats (95%CI: `r quantile(avg_change, c(0.025, 0.975))[1]` to `r quantile(avg_change, c(0.025, 0.975))[2]`%). Ectotherms in marine and freshwater environments showed partial compensation of physiological rates ([@fig-1]B) amounting to reduced $lnRR_{Q_{10_{acclimation}}}$ of `r mean(change_f)`% (95% CI: `r quantile(change_f, c(0.025, 0.975))[1]` to `r quantile(change_f, c(0.025, 0.975))[2]`) in freshwater and `r mean(change_m)`% (95% CI: `r quantile(change_m, c(0.025, 0.975))[1]` to `r quantile(change_m, c(0.025, 0.975))[2]`) in marine environments. In contrast, terrestrial ectotherms showed no acclimation (possibly even inverse acclimation) -- showing a `r mean(change_t)`% increase in $lnRR_{Q_{10_{acclimation}}}$ (95% CI: `r quantile(change_t, c(0.025, 0.975))[1]` to `r quantile(change_t, c(0.025, 0.975))[2]`, [@fig-1]B). While different trait categories showed different accclimation responses across habitat types, overall, they mirrored patterns seen overall (@fig-2). Acclimation capacity did not vary by life-stage given that there were no differences between $lnRR_{Q_{10_{acclimation}}}$ and $lnRR_{Q_{10_{acute}}}$ between adult and jeuveniles (@fig-3) (Adult-Jeuvenile (Acute): `r mean(acute_all)`, 95% CI: `r quantile(acute_all, 0.025)[1]` to `r quantile(acute_all, 0.975)[1]`, $p_{mcmc}$ = `r pmcmc(acute_all)`; Adult-Jeuvenile (Acclimation): `r mean(acclim_all)`, 95% CI: `r quantile(acclim_all, 0.025)[1]` to `r quantile(acclim_all, 0.975)[1]`, $p_{mcmc}$ = `r pmcmc(acclim_all)`).

@@

The opportunity for selection, as captured by $lnCVR_{Q_{10}}$, showed a decrease across all habitat types but were especially pronouced in terrestrial and marine ectotherms. Overall, there was a `r mean(t_cvr_mean_ef)`% (95% CI: `r quantile(t_cvr_mean_ef, 0.025)[1]` to `r quantile(t_cvr_mean_ef, 0.975)[1]`, $p_{mcmc}$ = `r pmcmc(t_cvr_mean)`) reduction in variation in physiological rates for terrestrial ectotherms and a `r mean(m_cvr_mean_ef)`% (95% CI: `r quantile(m_cvr_mean_ef, 0.025)[1]` to `r quantile(m_cvr_mean_ef, 0.975)[1]`, $p_{mcmc}$ = `r pmcmc(m_cvr_mean)`) reduction in variation for marine ectotherms when temperatures increase by 10$^\circ$C. In contrast, freshwater ectotherms exhibited a much smaller reduction in physiologiocal rate variance at high temperatures (`r mean(f_cvr_mean_ef)`%, 95% CI: `r quantile(f_cvr_mean_ef, 0.025)[1]` to `r quantile(f_cvr_mean_ef, 0.975)[1]`, $p_{mcmc}$ = `r pmcmc(f_cvr_mean)`).

@@

Thermal variability (i.e., $CV$) explained variation in acclimation capacity among terrestrial ectotherms, but not for marine or freshwater populations (@fig-4). Terrestrial ectotherm populations with a history of greater thermal variability were more likely to show greater capacity for thermal acclimation compared to populations with lower thermal variability (`r mean(slope_t)`, 95%CI: `r quantile(slope_t, 0.025)[1]` to `r quantile(slope_t, 0.975)[1]`, $p_{mcmc}$ = `r pmcmc(slope_t)`; $R^2$ = `r r2_t*100`%). In contrast, thermal predcitability did not explain much variation across marine, freshwater or terrestrial ectotherm populations (@fig-4).

>>>>

The final dataset included a total of `r f <- taxa_sum |> dplyr::filter(habitat == "f") |> dplyr::summarise(spp = sum(spp), k = sum(n)); f[1,1]` freshwater (fishes = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Actinopterygii") |> dplyr::summarise(sp = sum(spp))`, Molluscs = `r taxa_sum |> dplyr::filter(habitat == "f" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`, Amphibians = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Amphibia") |> dplyr::summarise(sp = sum(spp))`, Reptiles = `r taxa_sum |> dplyr::filter(habitat == "f" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))`, Arthropods = `r taxa_sum |> dplyr::filter(habitat == "f" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`, and a single Crustacean and Nematode species), `r m <- taxa_sum |> dplyr::filter(habitat == "m") |> dplyr::summarise(spp = sum(spp), k = sum(n)); m[1,1]` marine (fishes = `r taxa_sum |> dplyr::filter(habitat == "m" & class %in% c("Actinopterygii", "Agnatha","Chondrichthyes")) |> dplyr::summarise(sp = sum(spp))`, Annelids = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Annelida") |> dplyr::summarise(sp = sum(spp))`, Molluscs = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`, Echinoderms = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Echinodermata") |> dplyr::summarise(sp = sum(spp))`, Reptiles = `r taxa_sum |> dplyr::filter(habitat == "m" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))`, Arthropods = `r taxa_sum |> dplyr::filter(habitat == "m" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`, and a single Crustacean and Cnidarian species), and `r t <- taxa_sum |> dplyr::filter(habitat == "t") |> dplyr::summarise(spp = sum(spp), k = sum(n)); t[1,1]` terrestrial species (Annelids = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Annelida") |> dplyr::summarise(sp = sum(spp))`, Molluscs = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Mollusca") |> dplyr::summarise(sp = sum(spp))`, Arthropods = `r taxa_sum |> dplyr::filter(habitat == "t" & phylum == "Arthropoda") |> dplyr::summarise(sp = sum(spp))`, Reptiles = `r taxa_sum |> dplyr::filter(habitat == "t" & class == "Reptilia") |> dplyr::summarise(sp = sum(spp))` and Amphibians = `r taxa_sum |> dplyr::filter(habitat == "t" & class == "Amphibia") |> dplyr::summarise(sp = sum(spp))` along with a single Tardigrade species) (@fig-1 A). We had more data on acute thermal responses (n = `r trimmed <- data_long2 |> dplyr::filter(grepl("lnRR", name) & !is.na(effect_size) & !is.na(V) & !V == 0); sum1 = trimmed |> dplyr::group_by(name, type) |> dplyr::summarise(n = dplyr::n()) |> dplyr::filter(type =="acute"); sum(sum1$n)`) compared to thermal responses after an acclimation period (n = `r sum2 <- trimmed |> dplyr::group_by(name, type) |> dplyr::summarise(n = dplyr::n()) |> dplyr::filter(type =="acclim"); sum2$n`) because both acclimation temperatures had separate acute responses (@fig-1). While the two acute $lnRR_{Q_{10}}$ effect sizes did differ significantly from each other, on average (Acute responses were higher for animals acclimated to high temperatures -- $\beta$ = `r model1.s1$b[3]`, 95% CI: `r model1.s1$ci.lb[3]` to `r model1.s1$ci.ub[3]`, $p_{MCMC}$ = `r p_value(model1.s1$pval[3])`), they were in the same direction and only differed by ~10%. As such, we averaged the two acute $lnRR_{Q_{10}}$ effect sizes in subsequent analyses.

@@

Overall, $lnRR_{Q_{10_{acclimation}}}$ and $lnRR_{Q_{10_{acute}}}$ differed by only `r mean(avg_change)`% across all habitats (95%CI: `r quantile(avg_change, c(0.025, 0.975))[1]` to `r quantile(avg_change, c(0.025, 0.975))[2]`%). Ectotherms in marine and freshwater environments showed partial compensation of physiological rates ([@fig-1]B) amounting to reduced $lnRR_{Q_{10_{acclimation}}}$ of `r mean(change_f)`% (95% CI: `r quantile(change_f, c(0.025, 0.975))[1]` to `r quantile(change_f, c(0.025, 0.975))[2]`) in freshwater and `r mean(change_m)`% (95% CI: `r quantile(change_m, c(0.025, 0.975))[1]` to `r quantile(change_m, c(0.025, 0.975))[2]`) in marine environments. In contrast, terrestrial ectotherms showed no acclimation (possibly even inverse acclimation) -- showing a `r mean(change_t)`% increase in $lnRR_{Q_{10_{acclimation}}}$ (95% CI: `r quantile(change_t, c(0.025, 0.975))[1]` to `r quantile(change_t, c(0.025, 0.975))[2]`, [@fig-1]B). While different trait categories showed different accclimation responses across habitat types, overall, they mirrored patterns seen overall (@fig-2). Acclimation capacity did not vary by life-stage given that there were no differences between $lnRR_{Q_{10_{acclimation}}}$ and $lnRR_{Q_{10_{acute}}}$ between adult and jeuveniles (@fig-3) (Adult-Jeuvenile (Acute): `r mean(acute_all)`, 95% CI: `r quantile(acute_all, 0.025)[1]` to `r quantile(acute_all, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(acute_all)`; Adult-Jeuvenile (Acclimation): `r mean(acclim_all)`, 95% CI: `r quantile(acclim_all, 0.025)[1]` to `r quantile(acclim_all, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(acclim_all)`).

@@

The opportunity for selection, as captured by $lnCVR_{Q_{10}}$, showed a decrease across all habitat types but were especially pronouced in terrestrial and marine ectotherms. Overall, there was a `r mean(t_cvr_mean_ef)`% (95% CI: `r quantile(t_cvr_mean_ef, 0.025)[1]` to `r quantile(t_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(t_cvr_mean)`) reduction in variation in physiological rates for terrestrial ectotherms and a `r mean(m_cvr_mean_ef)`% (95% CI: `r quantile(m_cvr_mean_ef, 0.025)[1]` to `r quantile(m_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(m_cvr_mean)`) reduction in variation for marine ectotherms when temperatures increase by 10$^\circ$C. In contrast, freshwater ectotherms exhibited a much smaller reduction in physiologiocal rate variance at high temperatures (`r mean(f_cvr_mean_ef)`%, 95% CI: `r quantile(f_cvr_mean_ef, 0.025)[1]` to `r quantile(f_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(f_cvr_mean)`).

@@

Thermal variability (i.e., $CV$) explained variation in acclimation capacity among terrestrial ectotherms, but not for marine or freshwater populations (@fig-4). Terrestrial ectotherm populations with a history of greater thermal variability were more likely to show greater capacity for thermal acclimation compared to populations with lower thermal variability (`r mean(slope_t)`, 95%CI: `r quantile(slope_t, 0.025)[1]` to `r quantile(slope_t, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(slope_t)`; $R^2$ = `r r2_t*100`%). In contrast, thermal predcitability did not explain much variation across marine, freshwater or terrestrial ectotherm populations (@fig-4).
